---
title: "Explr_init"
output: html_document
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bumblebee data
A random sample of 25k was taken from the NBN data using pandas.dataframe.sample
This will be used to conduct an initial exploration of the data, and if successful, an initial attempt at creating an SDM for UK bumblebees.
This data was sourced from multiple organisations ('$ Data provider' in the data) which collected data on bumblebee occurrence independent to the NBN (check).

#Load library
```{r}
library(tidyverse)
library(ggplot2)
```

#Read dataset
```{r}
library(readr)
sampledata <- read_csv("nbn_bombus_occurence_sample_25k.csv")
```

#look at data
```{r}
str(sampledata)
glimpse(sampledata)
```

#Plot

```{r}
ggplot(sampledata, aes(x = `Start date`, y = `Scientific name`)) +
  geom_point()+ 
  ylab("Species") +
  xlab("Date") +
  theme_bw()
```

#We see...

```{r}
ggplot(data = sampledata, aes(x = `Scientific name`)) + 
  geom_bar() + 
  ylab("Frequency") +
  xlab("Species") +
  coord_flip() +
  theme_bw()
```

#Mapping the data

```{r}
library(sf)  # For spatial data handling
library(rnaturalearth)  # For base maps
library(rnaturalearthdata)  # Additional map data
library(dplyr)  # For data manipulation
```

```{r}
# Convert to spatial points using sf

bumblebee_sf <- sampledata %>%
  st_as_sf(coords = c("Longitude (WGS84)", "Latitude (WGS84)"), crs = 4326) %>%
  mutate(date = as.factor(sampledata$`Start date`))  # Ensure date is in the dataset

bumblebee_sf <- sampledata %>%
  st_as_sf(coords = c("Longitude (WGS84)", "Latitude (WGS84)"), crs = 4326)

# Ensure species is treated as a factor
bumblebee_sf$species <- as.factor(sampledata$`Scientific name`)

```


```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
uk <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "United Kingdom")
```

```{r}
ggplot() +
  geom_sf(data = uk, fill = "gray80", color = "black") +  # Base map
  geom_sf(data = bumblebee_sf, aes(color = species), size = 2, alpha = 0.7) +  # Plot occurrences
  scale_color_viridis_d() +  # Color scale for species
  theme_minimal() +
  labs(title = "Bumblebee Species Occurrence",
       x = "Longitude", 
       y = "Latitude", 
       color = "Species") +
  theme(legend.position = "bottom")


```








